---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
### Unzip the activity zip file
```{r}
unzip("activity.zip")
```
### Load the data into a variable
```{r}
raw_data <- read.csv("activity.csv")
```

### Convert the date column to R Date.
```{r}
raw_data$date <- as.Date(as.character(raw_data$date), "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
### Aggregate step data as a sum, for each day.
```{r}
steps_per_day <- aggregate(steps ~ date, FUN=sum, data=raw_data)
```
### Compute a histogram of daily step data.
```{r}
hist(steps_per_day$steps)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r}
avg <- mean(steps_per_day$steps)
med <- median(steps_per_day$steps)
```
Average daily step count: `r avg`.  
Median daily step count: `r med`.



## What is the average daily activity pattern?
### Compute an average of the number of steps per 5-minute interval
```{r}
daily_aggregate <- aggregate(steps ~ interval, FUN=sum, data=raw_data, na.action=na.omit)
plot(daily_aggregate$interval, daily_aggregate$steps, type = "l")
max(daily_aggregate$steps)

```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
